---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
---


## Loading and preprocessing the data


```{r,cache=TRUE}
library(data.table)
activityData <- read.csv("activity.csv")
populated <- read.csv("activity.csv")
naCases <- complete.cases(activityData)
cleanData <- activityData[naCases,]
```


## What is mean total number of steps taken per day?
```{r}
steps <-aggregate(cleanData$steps, list(date=cleanData$date),sum)
step_median <-median(steps$x,na.rm=TRUE)
step_mean <- mean(steps$x,na.rm=TRUE)
```
Mean total number steps per day = `r step_mean`

Median of total number of steps per day = `r step_median`

### Make a histogram of the total number of steps taken each day
```{r}
hist(steps$x,main="Distribution of Steps taken per day",
            xlab="Steps take per day",col="orange",breaks=seq(0,25000,1000))
```

### What is the average daily activity pattern?

```{r}
steps_interval <- aggregate(cleanData$steps, list(interval=cleanData$interval),mean)
    plot(steps_interval$interval,steps_interval$x,xlab="Time interval",type="l"
         ,ylab="No of steps taken",lwd=1,col='brown',
         main="Average daily activity pattern")
```

#####Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r echo=FALSE}
   max_step_interval = steps_interval[which.max(steps_interval$x),]
```
The maximum steps taken is in the interval `r max_step_interval$interval` and the max steps taken in this interval is `r max_step_interval$x`

### Imputing missing values

#####Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

    The number of observations/rows having NA's are `r sum(naCases == FALSE)`

#####Imputing the missing values
    The stratergy used to for imputing the missing values is replace NA's with the mean of the interval across the activity dataset.

```{r }
    for(i in 1:length(activityData$steps)){
        if(is.na(activityData[i,1])){
            value <- steps_interval[activityData[i,3] == steps_interval$interval,2]
            populated[i,1] = value
        }
    }
    populated_steps <-aggregate(populated$steps, list(date=populated$date),sum)

```
    
#####Created Data set
```{r}
summary(populated)
```

#####Check for missing values in the new data set
```{r}
sum(is.na(populated$steps))
```

###Make a histogram of the total number of steps taken each day with new data set
```{r}
    hist(populated_steps$x,main="Distribution of Steps taken per day with NA's imputed",
            xlab="Steps take per day",col="orange",breaks=seq(0,25000,1000))
```            

##### Calculate and report the mean and median


####Before Filling the data

Mean :
```{r echo=FALSE}
print(step_mean) 
```

Median :
```{r echo=FALSE}
print(step_median) 
```

```{r echo=FALSE}
    
popu_step_mean <- mean(populated_steps$x,na.rm=TRUE)
popu_step_median <- median(populated_steps$x,na.rm=TRUE)
```

####After Filling the Data:

Mean:
```{r echo=FALSE}
print(popu_step_median)
```

Median:
```{r echo=FALSE}
print(popu_step_mean)
```



There is no change in the mean's only there is a slight change in median's. We can conclude
that by populating the NA's with average of steps across the interval, only the increases the 
max values slightly



### Are there differences in activity patterns between weekdays and weekends?

####Calculating the aggregate of steps in weekdays and weekend
```{r }
    library(lubridate)
    populated$dayType <-0
   
    populated$date <- as.Date(populated$date)
    for(i in 1:length(populated$steps)){
    
        if(wday(populated$date[i]) %in% c(1,2,3,4,5)){
            populated$dayType[i]="weekday" 
        }
        else {
            populated$dayType[i]="weekend"
        }
    }
```


####Ploting the no of steps disbtribution between Weekdays and Weekends

```{r}
library(ggplot2) # we shall use ggplot2 for plotting figures
    averages <- aggregate(steps ~ interval + dayType, data=populated, mean)
    ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(dayType ~ .) +xlab("5-minute interval") +   ylab("Number of steps")

```

####Observation

The peaks for entire range of intervals go up for the weekend when compared to weekedays. On weekends we see a wider range of intervals having peaks greater than 100,activity spread over interval range is high.